---
title: "Getting started with RaCInG"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting started with RaCInG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`RaCInG` reconstructs patient-specific cell-cell communication networks from bulk
RNA-seq data. The package supports two complementary workflows:

- a **kernel-based** approach for fast deterministic feature extraction, and
- a **Monte Carlo** approach for simulation-based network summaries.

This vignette shows the recommended workflow and the most important entry points for
new users.

## Installation

### Install from GitHub

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("mhurtado13/racing")
library(RaCInG)
```

### Install from a local checkout

```{r eval = FALSE}
# install.packages("devtools")
devtools::install(".")
library(RaCInG)
```

If you want to build the RaCInG input matrices directly from raw counts, install the
optional preprocessing dependencies used by `prepare_input_files()`:

```{r eval = FALSE}
install.packages(c("ggplot2", "OmnipathR"))
# Additional optional packages: ADImpute, multideconv, liana
```

## Workflow at a glance

| Goal | Function | Output |
| --- | --- | --- |
| Build input matrices from raw counts | `prepare_input_files()` | Named list with `L`, `R`, `C`, `LR` matrices and labels |
| Compute deterministic features | `compute_racing_kernel()` | Kernel arrays + feature matrix |
| Compute simulation summaries | `compute_racing_montecarlo()` | Processed Monte Carlo results |
| Compare clinical groups | `wilcox_group_test()` | Statistics table and volcano plot |

## Recommended workflow

### 1. Build input matrices from raw counts

Use `prepare_input_files()` to generate, save, and load the input matrices in a
single call.

```{r eval = FALSE}
input <- prepare_input_files(
  counts = counts_matrix,
  output_folder = "Results/",
  file_name = "SKCM"
)

str(input)
```

### 2. Run the kernel method

The kernel method is the fastest way to derive direct, wedge, triangle, or GSCC
features across patients.
You can pass `counts` to let the function compute inputs
automatically, or supply previously computed matrices via `input_data`.

```{r eval = FALSE}
# Option A: from raw counts (runs prepare_input_files internally)
kernel_res <- compute_racing_kernel(
  counts = counts_matrix,
  output_folder = tempdir(),
  file_name = "SKCM",
  communication_type = "W",
  norm = TRUE
)

# Option B: from pre-computed input matrices (skips preprocessing)
kernel_res <- compute_racing_kernel(
  input_data = input,
  communication_type = "W",
  norm = TRUE
)

head(kernel_res$features[, 1:5])
```

### 3. Run the Monte Carlo method

Use the Monte Carlo workflow when you want simulation-based summaries or uncertainty
estimates from repeated graph realizations. The same `input_data` shortcut is
available here.

```{r eval = FALSE}
set.seed(1)
mc_res <- compute_racing_montecarlo(
  counts = counts_matrix,
  output_folder = tempdir(),
  deconv_method = "Quantiseq",
  file_name = "SKCM",
  nPatients = 3,
  communication_type = "W",
  Ncells = 100,
  Ngraphs = 10,
  Ndegree = 3,
  norm = TRUE
)

# Or from pre-computed inputs:
mc_res <- compute_racing_montecarlo(
  input_data = input,
  output_folder = tempdir(),
  file_name = "SKCM",
  communication_type = "W",
  Ncells = 100,
  Ngraphs = 10,
  Ndegree = 3,
  norm = TRUE
)

head(mc_res$output$mean[, 1:5])
```

### 4. Perform statistical testing

Once features are available in a patient-by-feature matrix, use the built-in Wilcoxon
workflow to compare groups.

```{r eval = FALSE}
grouping <- c("Responder", "Responder", "Non-responder", "Non-responder")
wilcox_results <- wilcox_group_test(kernel_res$features, grouping)
head(wilcox_results)
volcano_plot(wilcox_results, top_labels = 15)
```

## Notes

- `compute_racing_kernel()` and `compute_racing_montecarlo()` are the main entry points.
- Both accept an `input_data` argument with pre-computed matrices (as returned by
  `prepare_input_files()`), which skips all preprocessing.
- `prepare_input_files()` requires extra packages for deconvolution and prior-network
  assembly.
- The original Python implementation is available at
  <https://github.com/SysBioOncology/RaCInG>.

## Understanding the input files

RaCInG requires four matrices and associated label vectors that describe the
cell-cell communication landscape for a cohort of patients. The table below
summarises each component:

| Component | Dimensions | Description |
| --- | --- | --- |
| **Lmatrix** | cell types × ligands | Expression weight of each ligand in each cell type. Rows are cell types; columns are ligands. |
| **Rmatrix** | cell types × receptors | Expression weight of each receptor in each cell type. Same row order as `Lmatrix`. |
| **Cmatrix** | patients × cell types | Cell-type fraction for each patient. Each row sums to 1. |
| **LRmatrix** | ligands × receptors × patients | 3-D tensor of ligand–receptor interaction weights. Each patient slice is normalised to sum to 1. |
| **celltypes** | character vector | Alphabetically sorted cell-type names (shared across L, R, and C). |
| **ligands** | character vector | Ligand names matching the columns of `Lmatrix` and the first axis of `LRmatrix`. |
| **receptors** | character vector | Receptor names matching the columns of `Rmatrix` and the second axis of `LRmatrix`. |
| **Sign_matrix** | ligands × receptors | Optional matrix encoding known stimulatory (+1) or inhibitory (−1) interactions. Zeros indicate unknown. |

These inputs are typically generated by `prepare_input_files()` from a raw
counts matrix, or they can be assembled manually from pre-existing deconvolution
and prior-network data. The bundled `skcm_example` dataset provides a ready-made
example of this structure.

### Inspecting the example inputs

```{r}
library(RaCInG)
data(skcm_example)

# Overall structure
str(skcm_example, max.level = 1)
```

```{r}
# Lmatrix: 9 cell types × 276 ligands
dim(skcm_example$Lmatrix)
skcm_example$Lmatrix[1:4, 1:5]
```

```{r}
# Rmatrix: 9 cell types × 298 receptors
dim(skcm_example$Rmatrix)
skcm_example$Rmatrix[1:4, 1:5]
```

```{r}
# Cmatrix: 10 patients × 9 cell types (rows sum to 1)
dim(skcm_example$Cmatrix)
skcm_example$Cmatrix[1:4, ]
```

```{r}
# LRmatrix: 276 ligands × 298 receptors × 10 patients (3-D tensor)
dim(skcm_example$LRmatrix)
# First patient slice, top-left corner:
skcm_example$LRmatrix[1:4, 1:4, 1]
```

```{r}
# Label vectors
skcm_example$celltypes
head(skcm_example$ligands, 10)
head(skcm_example$receptors, 10)
```

## Running with the bundled example data

The `skcm_example` list shown above can be passed directly to the kernel or Monte
Carlo workflows via the `input_data` parameter.

### Kernel method on the example data

```{r eval = FALSE}
kernel_res <- compute_racing_kernel(
  input_data   = skcm_example,
  output_folder = tempdir(),
  communication_type = "W",
  norm = TRUE
)

head(kernel_res$features[, 1:5])
```

### Monte Carlo method on the example data

```{r eval = FALSE}
set.seed(42)
mc_res <- compute_racing_montecarlo(
  input_data   = skcm_example,
  output_folder = tempdir(),
  file_name     = "skcm_example",
  nPatients     = 2,
  communication_type = "W",
  Ncells  = 100,
  Ngraphs = 5,
  Ndegree = 3,
  norm    = TRUE
)
```


