---
title: "Getting started with RaCInG"
author: "Marcelo Hurtado"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting started with RaCInG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`RaCInG` reconstructs patient-specific cell-cell communication networks from bulk
RNA-seq data. The package supports two complementary workflows:

- a **kernel-based** approach for fast deterministic feature extraction, and
- a **Monte Carlo** approach for simulation-based network summaries.

This vignette shows the recommended workflow and the most important entry points for
new users.

## Installation

### Install from GitHub

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("mhurtado13/racing")
library(RaCInG)
```

### Install from a local checkout

```{r eval = FALSE}
# install.packages("devtools")
devtools::install(".")
library(RaCInG)
```

If you want to build the RaCInG input matrices directly from raw counts, install the
optional preprocessing dependencies used by `prepare_input_files()`:

```{r eval = FALSE}
install.packages(c("ggplot2", "OmnipathR"))
# Additional optional packages: ADImpute, multideconv, liana
```

## Workflow at a glance

| Goal | Function | Output |
| --- | --- | --- |
| Build input matrices from raw counts | `prepare_input_files()` | Named list with `L`, `R`, `C`, `LR` matrices and labels |
| Compute deterministic features | `compute_racing_kernel()` | Kernel arrays + feature matrix |
| Compute simulation summaries | `compute_racing_montecarlo()` | Processed Monte Carlo results |
| Compare clinical groups | `wilcox_group_test()` | Statistics table and volcano plot |

## Recommended workflow

### 1. Build input matrices from raw counts

Use `prepare_input_files()` to generate, save, and load the input matrices in a
single call.

```{r eval = FALSE}
input <- prepare_input_files(
  counts = counts_matrix,
  output_folder = "Results/",
  file_name = "SKCM"
)

str(input)
```

### 2. Run the kernel method

The kernel method is the fastest way to derive direct, wedge, triangle, or GSCC
features across patients.
You can pass `counts` to let the function compute inputs
automatically, or supply previously computed matrices via `input_data`.

```{r eval = FALSE}
# Option A: from raw counts (runs prepare_input_files internally)
kernel_res <- compute_racing_kernel(
  counts = counts_matrix,
  output_folder = tempdir(),
  file_name = "SKCM",
  communication_type = "W",
  norm = TRUE
)

# Option B: from pre-computed input matrices (skips preprocessing)
kernel_res <- compute_racing_kernel(
  input_data = input,
  communication_type = "W",
  norm = TRUE
)

head(kernel_res$features[, 1:5])
```

### 3. Run the Monte Carlo method

Use the Monte Carlo workflow when you want simulation-based summaries or uncertainty
estimates from repeated graph realizations. The same `input_data` shortcut is
available here.

```{r eval = FALSE}
set.seed(1)
mc_res <- compute_racing_montecarlo(
  counts = counts_matrix,
  output_folder = tempdir(),
  deconv_method = "Quantiseq",
  file_name = "SKCM",
  nPatients = 3,
  communication_type = "W",
  Ncells = 100,
  Ngraphs = 10,
  Ndegree = 3,
  norm = TRUE
)

# Or from pre-computed inputs:
mc_res <- compute_racing_montecarlo(
  input_data = input,
  output_folder = tempdir(),
  file_name = "SKCM",
  communication_type = "W",
  Ncells = 100,
  Ngraphs = 10,
  Ndegree = 3,
  norm = TRUE
)

head(mc_res$output$mean[, 1:5])
```

### 4. Perform statistical testing

Once features are available in a patient-by-feature matrix, use the built-in Wilcoxon
workflow to compare groups.

```{r eval = FALSE}
grouping <- c("Responder", "Responder", "Non-responder", "Non-responder")
wilcox_results <- wilcox_group_test(kernel_res$features, grouping)
head(wilcox_results)
volcano_plot(wilcox_results, top_labels = 15)
```

## Notes

- `compute_racing_kernel()` and `compute_racing_montecarlo()` are the main entry points.
- Both accept an `input_data` argument with pre-computed matrices (as returned by
  `prepare_input_files()`), which skips all preprocessing.
- `prepare_input_files()` requires extra packages for deconvolution and prior-network
  assembly.
- The original Python implementation is available at
  <https://github.com/SysBioOncology/RaCInG>.

## Running with the bundled example data

The package ships with `skcm_example`, a pre-processed named list derived from 10
TCGA SKCM melanoma patients. It can be passed directly to the kernel or Monte Carlo
workflows via the `input_data` parameter.

```{r}
library(RaCInG)
data(skcm_example)
str(skcm_example, max.level = 1)
```

### Kernel method on the example data

```{r eval = FALSE}
kernel_res <- compute_racing_kernel(
  input_data   = skcm_example,
  output_folder = tempdir(),
  communication_type = "W",
  norm = TRUE
)

head(kernel_res$features[, 1:5])
```

### Monte Carlo method on the example data

```{r eval = FALSE}
set.seed(42)
mc_res <- compute_racing_montecarlo(
  input_data   = skcm_example,
  output_folder = tempdir(),
  file_name     = "skcm_example",
  nPatients     = 2,
  communication_type = "W",
  Ncells  = 100,
  Ngraphs = 5,
  Ndegree = 3,
  norm    = TRUE
)
```


