---
title: "Getting started with RaCInG"
author: "Marcelo Hurtado"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting started with RaCInG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`RaCInG` reconstructs patient-specific cell-cell communication networks from bulk
RNA-seq data. The package supports two complementary workflows:

- a **kernel-based** approach for fast deterministic feature extraction, and
- a **Monte Carlo** approach for simulation-based network summaries.

This vignette shows the recommended workflow and the most important entry points for
new users.

## Installation

### Install from GitHub

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("mhurtado13/racing")
library(RaCInG)
```

### Install from a local checkout

```{r eval = FALSE}
# install.packages("devtools")
devtools::install(".")
library(RaCInG)
```

If you want to build the RaCInG input matrices directly from raw counts, install the
optional preprocessing dependencies used by `prepare_input_files()`:

```{r eval = FALSE}
install.packages(c("ggplot2", "OmnipathR"))
# Additional optional packages: ADImpute, multideconv, liana
```

## Workflow at a glance

| Goal | Function | Output |
| --- | --- | --- |
| Build input files from raw counts | `prepare_input_files()` | `L`, `R`, `C`, and `LR` matrices |
| Load prepared inputs | `generateInput()` | Named list of matrices and labels |
| Compute deterministic features | `compute_racing_kernel()` | Kernel arrays + feature matrix |
| Compute simulation summaries | `compute_racing_montecarlo()` | Processed Monte Carlo results |
| Compare clinical groups | `wilcox_group_test()` | Statistics table and volcano plot |

## Recommended workflow

### 1. Start from prepared input matrices

If you already have precomputed `L`, `R`, `C`, and `LR` matrices, load them with
`generateInput()`.

```{r eval = FALSE}
input <- generateInput(
  file_name = "SKCM",
  output_folder = "Results/"
)

str(input)
```

### 2. Run the kernel method

The kernel method is the fastest way to derive direct, wedge, triangle, or GSCC
features across patients.

```{r eval = FALSE}
kernel_res <- compute_racing_kernel(
  counts = counts_matrix,
  output_folder = tempdir(),
  file_name = "SKCM",
  communication_type = "W",
  norm = TRUE
)

head(kernel_res$features[, 1:5])
```

### 3. Run the Monte Carlo method

Use the Monte Carlo workflow when you want simulation-based summaries or uncertainty
estimates from repeated graph realizations.

```{r eval = FALSE}
set.seed(1)
mc_res <- compute_racing_montecarlo(
  counts = counts_matrix,
  output_folder = tempdir(),
  deconv_method = "Quantiseq",
  file_name = "SKCM",
  nPatients = 3,
  communication_type = "W",
  Ncells = 100,
  Ngraphs = 10,
  Ndegree = 3,
  norm = TRUE
)

head(mc_res$output$mean[, 1:5])
```

### 4. Perform statistical testing

Once features are available in a patient-by-feature matrix, use the built-in Wilcoxon
workflow to compare groups.

```{r eval = FALSE}
grouping <- c("Responder", "Responder", "Non-responder", "Non-responder")
wilcox_results <- wilcox_group_test(kernel_res$features, grouping)
head(wilcox_results)
volcano_plot(wilcox_results, top_labels = 15)
```

## Notes

- `compute_racing_kernel()` and `compute_racing_montecarlo()` are the main entry points.
- `prepare_input_files()` requires extra packages for deconvolution and prior-network
  assembly.
- The original Python implementation is available at
  <https://github.com/SysBioOncology/RaCInG>.


